---
title: "endpointR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{endpointR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## About endpointR
In our paper **Body weight course as a reliable marker for humane endpoint determination in rat models with intracranial tumor** by Helgers and Talbot et al (2019) we describe an algorithm that is capable of detecting endpoints in different experimental data sets with a detection rate of 97 \%. As the algorithm is largely based on the detection of variance deviation, its main flaw is false alarms. However, as a tool for the unexperienced researcher doing research with laboratory animals it is a valuable addition to their daily work. Endpoint detection requires a high level of training and may not be objective.

With the presented algorithm a help in the decision making process was offered, so that in critical situations researchers still have the option of rating the animals in a more traditional way. Another finding was the high correlation of clinical scoring with the mean absolute deviation (MAD) of the body weight data. By constraining the body weight variance space by MAD almost similar results as with empirical clinical scoring was achieved. 

Therefore, we present the endpointR algorithm in this package, so that it can be used with body weight data in other models or even with other variables for potential humane endpoint determination.

## What can it do - what is it for?



## Functions

## working example



```{r setup}
library(endpointR)
```
